F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-24 icfp-2015/paper.tex
Finishing the proofs and discussions

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}

@@

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
% TODO: Cleanup "strucutred" vs "structural"
% TODO: Call it "relativized safety" or "relativized soundness" consistently

@@

\vspace{0.5em}
\noindent\boxed{\emph{Primitives, options, collections}}
\begin{align}
\label{eq:sub-prim}\tag{B1}
\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\quad\\[-0.2em]
\label{eq:sub-null}\tag{B2}
\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
\label{eq:sub-opt}\tag{B3}
\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
\label{eq:sub-top}\tag{B4}
\sigma :> \top \quad &(\textnormal{for all}~\sigma)\\[-0.2em]
\label{eq:sub-col}\tag{B5}
[\sigma_1] :> [\sigma_2] \quad &(\textnormal{if}~\sigma_1 :> \sigma_2)
\end{align}

\noindent\boxed{\emph{Union types}}
\begin{align}
\label{eq:sub-union1}\tag{U1}
\sigma_1 + \ldots + \sigma_n :> \sigma_1' + \ldots + \sigma_n' \quad &(\textnormal{iff}~\forall i.\sigma_i :> \sigma_i') \\
\label{eq:sub-union2}\tag{U2}
\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \leq n) \\
\label{eq:sub-union3}\tag{U3}
\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
\label{eq:sub-union-el}\tag{U4}
\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
\end{align}

@@

\begin{figure*}[t]
\begin{equation*}
\hspace{3em}
\inference[(record-1)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \nu \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \nu \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

@@

% all rules are symmetric and reflexive
\begin{equation*}
\hspace{4.75em}
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\qquad\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.75em]
 % top type
 \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
\end{array}
%
\qquad\quad 
% null and nullable
 \begin{array}{l}
 \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.75em]
 \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
 \end{array}
\end{equation*}

@@

\item The remaining rules for unins are standard. Subtype can have fewer alternatives 
  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
  (\ref{eq:sub-union3}). Similarly, the collection type is also covariant (\ref{eq:sub-col}).

@@

\caption{Featherweight F\# -- Remaining reduction rules}
\label{fig:ff-reduction}
\vspace{-0.5em}
\end{figure}

@@

Next, we define the reduction relation and (a fragment of) type checking for Featherweight F\#.
The language presented here is intentionally incomplete. We only define parts needed to prove
the relativized soundness property in Section~\ref{sec:safety}.

@@

\noindent
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep [\cdot] 
 \\[0.6em]
 f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \;\}
\end{array} 
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\qquad
\inference
  {~}
  {\Gamma \vdash s : \ident{StructVal}}
\qquad  
\inference
  {~}
  {\Gamma \vdash n : \ident{int}}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..}
  {\Gamma \vdash e.N_i:\tau_i}
\end{equation*}

\begin{equation*}
\inference
  {\Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots}
  {\Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{equation*}

@@

\paragraph{Type checking.} 
Selected type checking rules for Featherweight F\# are shown in Figure~\ref{fig:ff-typecheck}.
The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
%
\begin{itemize}
\item All structured values $s$ have a type \ident{StructVal}. Some of those have other types
  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
\item A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
  non-structural values like objects or options are not of type \ident{StructVal}.
\item The two rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
relativized safety theorem, which considers only classes generated by our type provider (which 
are correct by construction).

@@

\paragraph{Example 1.}
To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
that the inferred type is a record with two fields (one optional) such as 
$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
rules from Figure~\ref{fig:tp-generation} produces the following class:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
 \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
\end{array}
\end{equation*}
%
The body of the \ident{Age} member is produced by the case for optional types applied to an expression
$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
perform any checks.

For completeness, a type corresponding to the record is \ident{Person} and given a 
structural document $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.

@@

Now that we have

\begin{theorem}[Relativized safety]
Given samples $s_1, \ldots, s_n$, inferred type $\sigma=\semalt{s_1, \ldots, s_n}$ and generated 
$\tau,e,L = \sem{\sigma}_x$ then:

For all $s'$ such that $\exists i.s_i :> s'$, then for any $e$ such that $x:\ident{StructVal} \vdash e:\tau$
it is the case that $e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.

\end{theorem}
% - Safety property

@@

\subsection{Discussion}
\label{sec:safety-discuss}
alternative - check the whole file upfront based on the inferred type
what if the sample is online??

@@

\label{sec:impl-parsing}

Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)

Error handling - solve issues just add another example

\label{sec:impl-unions}
Unions become records with fields - so having null makes sense

@@

\label{sec:impl-collections}

@@

\newpage
~
\newpage

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

@@

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

>>>>

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}

@@

% TODO: The model here does not actually use type erasure!!
% TODO: Maybe get rid of \mu_opt for records everywhere
% TODO: Disambigurate 's' as structural value and 's' as string
% TODO: Explain why we can only provide simple classes (!)
% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
% TODO: Cleanup "strucutred" vs "structural"
% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)

@@

\vspace{0.5em}
\noindent\boxed{\emph{Primitives, options, collections}}
\begin{align}
\label{eq:sub-prim}\tag{B1}
\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\\[-0.2em]
\label{eq:sub-null}\tag{B2}
\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
\label{eq:sub-opt}\tag{B3}
\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
\label{eq:sub-top}\tag{B4}
\sigma :> \top \quad &(\textnormal{for all}~\sigma)\\[-0.2em]
\label{eq:sub-col}\tag{B5}
[\sigma_1] :> [\sigma_2] \quad &(\textnormal{if}~\sigma_1 :> \sigma_2)
\end{align}

\noindent\boxed{\emph{Union types}}
\begin{align}
\label{eq:sub-union1}\tag{U1}
\sigma_1 + \ldots + \sigma_n :> \sigma_1' + \ldots + \sigma_n' \quad &(\textnormal{iff}~\forall i.\sigma_i :> \sigma_i') \\
\label{eq:sub-union2}\tag{U2}
\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \leq n) \\
\label{eq:sub-union3}\tag{U3}
\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \geq n) \\
\label{eq:sub-union4}\tag{U4}
\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
\label{eq:sub-union-el}\tag{U5}
\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
\end{align}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\hspace{3em}
\inference[(record-1)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \nu \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \nu \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

@@

% all rules are symmetric and reflexive
\noindent
\begin{equation*}
\hspace{4.75em}
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\qquad\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
 % top type
 \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
\end{array}
%
\qquad\quad 
% null and nullable
 \begin{array}{l}
 \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.6em]
 \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
 \end{array}
\end{equation*}

@@

\item For unions, we include the standard rules. Subtype can have fewer alternatives 
  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
  (\ref{eq:sub-union4}). An unusual rule (\ref{eq:sub-union3}) says that subtype can also have 
  \emph{more} cases. This is valid because the type provider exposes unions as objects that do not
  allow complete pattern matching.

@@

\caption{Featherweight F\# -- Remaining reduction rules}
\label{fig:ff-reduction}
\vspace{-1em}
\end{figure}

@@

Next, we define the reduction relation and (a fragment of) type checking for Featherweight F\#.
The language presented here is intentionally incomplete. We only define parts needed to prove
the relativized safety property in Section~\ref{sec:safety}.

@@

\vspace{0.5em}
\noindent
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep [\cdot] 
 \\[0.6em]
 f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \;\}
\end{array} 
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\qquad
\inference
  {~}
  {L; \Gamma \vdash s : \ident{StructVal}}
\qquad  
\inference
  {~}
  {L; \Gamma \vdash n : \ident{int}}
\end{equation*}

\begin{equation*}
\inference
  {L; \Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}

\begin{equation*}
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{equation*}

@@

\paragraph{Type checking.} 
The type checking rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
%
\begin{itemize}[noitemsep]
\item[--] All structured values $s$ have a type \ident{StructVal}. Some of those have other types
  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
  non-structural values like objects or options are not of type \ident{StructVal}.
\item[--] Primitive operations $op$ are treated as functions, accepting \ident{StructVal} and producing
  an appropriate type as the result.
\item[--] The two rules for checking class construction and member access are similar to corresponding
  rules of Featherweight Java.  
\end{itemize}
%
An important part of Featherweight Java that is omitted here is the checking of type declarations
(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
relativized safety theorem, which considers only classes generated by our type provider (which 
are correct by construction).

@@

\paragraph{Example 1.}
To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
that the inferred type is a record with two fields (one optional) such as 
$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
rules from Figure~\ref{fig:tp-generation} produces the following class:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
 \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
\end{array}
\end{equation*}
%
The body of the \ident{Age} member is produced by the case for optional types applied to an expression
$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
perform any checks. 
For completeness, a type corresponding to the record is \ident{Person} and given a 
structural document $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.

@@

Informally, the safety property of F\# Data type providers states that, given representative sample
documents, any code that can be written using the provided types is guaranteed to work. We call this 
\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, the user can 
always use the type provider with an input that has a different structure than any of the samples -- and in 
this case, it is expected that the code will fail at runtime (throw an exception in the actual 
implementation or by getting stuck in our model).

The key question is, what is a representative sample? Given a set of sample documents, 
the provided type is guaranteed to work if the inferred type of the input is a subtype of any of the
sample documents. Going back to Section~\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}[noitemsep]
\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
\item[--] Records in the actual input can have additional fields
\item[--] Records in the actual input can have fewer fileds, provided that the type of the fields is marked as optional in the sample
\item[--] Union types in the input can have both fewer or more cases
\item[--] When we have a union type in the sample, the actual input can also contain just values of one of the union cases
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on inputs $s'$ that is a subtype of the sample $s$. More formally, we require that the 
provided expression (using $s'$ as the input) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

\begin{lemma}[Correctness of provided types]
\label{thm:tp-correctness}
Given a sample value $s$ and an input value $s'$ such that $\semalt{s} :> \semalt{s'}$
and provided type, expression and class declarations $\tau, e, L = \sem{\semalt{s}}_{s'}$, 
then $e \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
class $C$, it holds that $e.N_i \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the structure of $\sem{-}_e$. For primitives, the conversion functions accept all subtypes.
For other cases, analyse the the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, when \ident{isTag} returns
\ident{true}, then a value has the structure required by the corresponding case).
\end{proof}

\noindent
Now that we know that the provided types are correct with respect to the subtyping relation, we can
look at the main theorem of the paper. It statest that, for any input (which is a subtype of any of the
samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.

Using the standard approach to syntactic type safety [?], we prove the type preservation 
(reduction does not change type) and progress (an expression that is not a value can be reduced).

\begin{theorem}[Relativized safety]
\label{thm:safety}
Assume $s_1, \ldots, s_n$ are samples, $\sigma=\semalt{s_1, \ldots, s_n}$ is an inferred
type and $\tau,e,L = \sem{\sigma}_x$ are a type, expression and class definitions generated by a 
type provider.

Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and that does not contain any structural values
as sub-expressions, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
also $\emptyset; \vdash v : \tau$.
\end{theorem}
\begin{proof}
We discuss the two parts of the proof separately as type prservation (Lemma~\ref{thm:rs-preservation})
and progress (Lemma~\ref{thm:rs-progress}).
\end{proof}

\begin{lemma}[Preservation]
\label{thm:rs-preservation}
Given the class definitions $L$ generated by a type provider as specified in
the assumptions of Theorem~\ref{thm:safety}, then if $\Gamma \vdash e : \tau$ and 
$e \reduce^{*} e'$ then $\Gamma \vdash e' : \tau$.
\end{lemma}
\begin{proof}
By induction over the reduction $\reduce$. The cases for the ML subset of Featherweight F\# 
are standard. For (\emph{member}), we check that code generated by type providers
in Figure~\ref{fig:tp-generation} is well-typed.
\end{proof}

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the ML subset and object-oriented subset of the calculus. The problematic
part are the conversion functions (Figure~\ref{fig:op-conversions}). Given a structural value 
(which has a type \ident{StructVal} in our language), the reduction can get stuck if the value does
not have a structure required by a specific conversion function used.

The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
we carefuly state that we only consider expressions $e_c$ which ``[do] not contain any structural 
values as sub-expressions''. This makes sure that the only code working with structural values is the
code generated by the type provider.

\begin{lemma}[Progress]
\label{thm:rs-progress}
Given the assumptions and definitions from Theorem~\ref{thm:safety}, it is the case that
$e_c[y\leftarrow e_s] \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the typing derivation of $L; \emptyset \vdash e_c[y\leftarrow e_s] : \tau'$. 
The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.
\end{proof}

\subsection{Discussion}
\label{sec:safety-discuss}

The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
for programming langauges without type providers. However, it reflects the reality of programming with external
data sources that is increasingly important in the age of web [?]. So, type providers do not reduce the safety
-- they simply make the existing issues visible.

The actual implementation in F\# Data throws an exception for invalid inputs. In contrast, the calculus
presented here simply gets stuck. We could extend the calculus with exceptions, but that would obscure the 
purpose -- precisely specifying when the code using type providers does not ``go wrong.''

As mentioned earlier, \emph{relativized safety} specifies a sufficient, but not a necessary condition.
This is also reflected in the conversion functions which allow additional conversions not required by 
the subtyping relation. In practice, the additional flexibility proves useful (as minor variations in 
input formats are common). However, an interesting stricter alternative would be to check that an input 
has the required type when it is loaded -- and throw an exception immediately when loading data rather than 
on member access.

Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
type providers is that the types can be generated lazilly -- only classes that are actually used in the type
checked code are generated. Extending our formalism to capture this aspect is an interesting future work
relevant to other type providers -- including the World Bank type provider [?] that is also available
in the F\# Data library.

@@

@@

\subsection{Parsing}
\label{sec:impl-parsing}
Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)

\subsection{Unions}
\label{sec:impl-unions}
Mainly because type providers can only provide objects

Unions become records with fields - so hiding null makes sense
also it means the open-world semantics

\subsection{Heterogeneous collections}
\label{sec:impl-collections}

\subsection{Practical experience}
Error handling - solve issues just add another example

\section{Related and future work}

~

@@

~

F\# and Agda type providers

Kathleen Fisher's work

JSON inference stuff

LINQ and XML stuff

@@

\acks
Acknowledgments, if needed.

@@
